---
title: "STAT497 Homework-2"
author: "Batuhan SAYLAM ID:2429264"
date: "2023-11-29"
output:
  html_document: default
  pdf_document: default
---
Q2) \
a) \

```{r}
library(AER)
data("USMacroSWM")
data=na.omit(USMacroSWM[,"production"])
```
```{r}

plot(data)

```
```{r}
acf(data, na.action = na.pass)

```
```{r}
pacf(data, na.action = na.pass)

```
\
\
The first plot shows us that the process is not stationary since there exists trend line. 
ACF shows slow decay so this is indicator of nonstationary.
Partial ACF show us that only 1st lag is significant in PACF. 
\
\
\
b) 
\
```{r}
library(forecast)
lambda <- BoxCox.lambda(data)
trans<-BoxCox(data,lambda)
plot(trans)
```
```{r}

acf(trans, na.action = na.pass)
```

```{r}
pacf(trans, na.action = na.pass)
```
\
\
The first plot shows us that the process is not stationary since there exists trend line. 
ACF shows slow decay so this is indicator of nonstationary.
Partial ACF show us that only 1st lag is significant in PACF.

\
\
c) \
```{r}
library(pdR)
out<-HEGY.test(wts=trans, itsd=c(1,1,0), regvar=0, selectlags=list(mode="signf", Pmax=NULL))
out$stats
```
\
\
In this output, we will use p value of tpi_1 for regular unit root and use the p value of Fpi_11:12 for testing seasonal unit root since we have monthly data. Also, the system has a regular unit root since p-value of tpi_1 is greater than 0.05. However, does not have a seasonal unit root since p-value of Fpi_11:12 is less than 0.05. 
\
\


```{r}
out<-HEGY.test(wts=diff(trans), itsd=c(1,1,0), regvar=0, selectlags=list(mode="signf", Pmax=NULL))
out$stats
```
\
\
The differenced system has no regular unit root since p-value of tpi_1 is less than 0.05.
\
\
\
d)
\

```{r}

acf(diff(trans), na.action = na.pass)
```

```{r}
pacf(diff(trans), na.action = na.pass)

```
\
SARIMA(3,1,0)(1,0,0)[12] or SARIMA(3,1,4)(1,0,0)[12] 
\



Q3)

a)\
```{r}
data("UKgas")
data=na.omit(UKgas)
plot(data)
```

```{r}
acf(data, na.action = na.pass)
```

```{r}
pacf(data, na.action = na.pass)
```
\
\
The first plot shows us that the process is not stationary since there exists trend line and seasonal behaviour. 
ACF shows linear decay on seasonal lags (4,8,12,16,...) so this is indicator of nonstationary.
Partial ACF show us that only 4th lag is significant in PACF. 
\
\
b)\
```{r}
lambda <- BoxCox.lambda(data)
trans<-BoxCox(data,lambda)
plot(trans)
```

```{r}

acf(trans, na.action = na.pass)
```

```{r}
pacf(trans, na.action = na.pass)
```
\
\
The first plot shows us that the process is not stationary since there exists trend line and seasonal behaviour. 
ACF shows linear decay on seasonal lags (4,8,12,16,...) so this is indicator of nonstationary.
Partial ACF show us that only 1st lag is significant in PACF.
\
\
\
c) \
```{r}
library(pdR)
out<-HEGY.test(wts=trans, itsd=c(1,1,0), regvar=0, selectlags=list(mode="signf", Pmax=NULL))
out$stats
```
\
\
In this output, we will use p value of tpi_1 for regular unit root and use the p value of Fpi_3:4 for testing seasonal unit root since we have quarterly data. Also, the system has a regular unit root since p-value of tpi_1 is greater than 0.05. Also, it have a seasonal unit root since p-value of Fpi_3:4 is greater than 0.05. 
\
\

```{r}
library(forecast)
ndiffs(trans)
nsdiffs(diff((trans)))
```
\
Thus, I need to take one regular and one seasonal difference.
\

```{r}
newdata=diff(trans)

```
\
```{r}
out<-HEGY.test(wts=newdata, itsd=c(1,0,0), regvar=0, selectlags=list(mode="signf", Pmax=NULL))
out$stats
```
\
\
The differenced system has no regular unit root  since p-value of tpi_1 is less than 0.05 but the differenced system has seasonal unit root  since p-value of Fpi_3:4  is greater than 0.05.
\
\

```{r}
newdata=diff(newdata,4)
```




```{r}
out<-HEGY.test(wts=newdata, itsd=c(0,0,0), regvar=0, selectlags=list(mode="signf", Pmax=NULL))
out$stats
```
\
\
The differenced system has no regular unit root and seasonal unit root since p-value of tpi_1 and Fpi_3:4 are less than 0.05.
\
\
\
d)
\
```{r}

acf(newdata, na.action = na.pass)
```

```{r}
pacf(newdata, na.action = na.pass)

```
\
SARIMA(2,1,1)(1,1,0)[4]
\



